Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}

### 2017-08-18 revision/source/ident_manu.tex
revision resubmitted

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Wed Aug 16 18:18:02 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/b081f900ad818adbbf496e0b8888820e38bda065}{b081f900ad818adbbf496e0b8888820e38bda065}\newline Acronyms: \acfi{chla}; \acfi{cgem}; \acfi{do}; \acfi{dom}; \acfi{gom}; \acfi{lcs}; \acfi{marb}; \acfi{pom}; \acfi{rmse}; \acfi{zerod}}

@@

Estimates of oxygen production from closed bottle experiments were used to calibrate the \ac{zerod} model.  Surface water was collected from a fixed location in Pensacola Bay, Florida (30.35$^{\circ}$ N, 87.20$^{\circ}$ W) on September 25\textsuperscript{th}, 2013 (full details in \citetalias{MurrellIR}).  Water samples were placed into closed 100 mL glass bottles at the time of collection and transported back to the lab for treatments.  Concentrations of \ac{do} and nutrients (NH$_4^+$, NO$_2^-$, NO$_3^-$, PO$_4^{3-}$, and SiO$_3^{2-}$) were extracted from the samples within one hour of collection. Oxygen concentrations were established using Winkler reagents \citep{Parsons85} and titration using a MetrOhm Titrando with thiosulfate titrant calibrated against an iodate standard and electrochemical endpoint detection.  Ammonium was analyzed using a fluorometric method \citep{Holmes99}; other nutrients were analyzed using either a Thermo Fisher Aquakem 200 discrete analyzer or an Astoria-Pacific continuous flow analyzer using standard colorimetric methods \citepalias{APHA05}.

Oxygen production after an approximate 12 hour period (morning to evening) was estimated within the bottles using light and dark treaments.  All treatments were exposed to ambient outdoor conditions in a water bath with continuous flow at the laboratory.  The dark treatments used opaque black bottles to ensure no exposure of ambient light to the samples.  The exposure experiments began at 7am and were concluded at 7pm, after which the concentrations of \ac{do} in each bottle were measured.  Each treatment had four replicates and HOBO\textsuperscript{\textregistered} pendants were used to measure continuous temperature and light.

Each parameter subset in \cref{tab:heurist1,tab:heurist2,tab:heurist3} for \ac{do} was calibrated to the ending oxygen concentration for each of the four replicates of each treatment. The intial nutrient and oxygen concentrations were used as starting conditions for FishTank.  The corresponding continuous temperature and light measurements were also used as forcing conditions.  Parameter values were searched using the optim function in R that implemented the limited memory modification of the BFGS quasi-Newton method \citep{Byrd95,Nocedal06,RDCT17}.  This optimization scheme searches for values within the user-defined lower and upper bounds of each parameter.  We set the algorithm to search within a range of values for a subset that was chosen to best reflect expectations given professional judgment and reasonable bounds from the literature.  While we recognize that `actual' parameter values could possibly extend beyond these ranges, constraining the values was expected to produce output that was more interpretable within the context of the sensitivity functions defined above.  Moreover, precise limits on each parameter were less of a concern for the current analysis as our primary goal was to demonstrate differences in fit as a function of sensitivity and identifiability.  The optimal values for each parameter were identified based on convergence of the model results as a minimization of the \ac{rmse} between the observed and modelled data.

@@

An evaluation of the effects of individual parameters on $\gamma$ suggested that some parameters had disproportionate effects on identifiability.  Based on $\gamma = 15$, \cref{fig:identplo} suggests that most parameter sets for organic matter were identifiable, regardless of how many parameters were selected (i.e., two through eight).  However, some subsets were not identifiable such that identification of one or more redundant parameters that are inflating $\gamma$ values could provide useful information.  \Cref{fig:exclex} shows an alternative view of identifiability of \ac{do} with exclusion and inclusion of individual parameters in different sets for the organic matter category.  As before, collinearity increases with more parameters in a subset, although the increase varies depending on which parameter was included or excluded from the set.  For example, inclusion of \textit{KNO3} in a parameter set almost always inflated $\gamma$.  All parameter subsets that did not include \textit{KNO3} were well below $\gamma = 15$, suggesting that exclusion of this parameter improves identifiability.  Interestingly, the inclusion of some parameters caused a reduction in $\gamma$, which contradicts the general rule that more parameters caused reduced identifiablity.  For example, parameter sets that included \textit{KGcdom} generally had lower $\gamma$ values relative to those that excluded the parameter.

@@

Calibration of parameter values for each of the seven selection heuristics (\cref{tab:heurist1,tab:heurist2,tab:heurist3}) produced model output that varied in goodness of fit and computational effort required to identify optimal values (\cref{tab:calib}).  As expected, the number of required iterations of the optimization algorithm to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only 12 iterations were required to optimize the single parameter selected for the temperature category, whereas 615 iterations were required for the seven parameters selected for phytoplankton (\cref{fig:calex}).  No association was observed between collinearity ($\gamma$) and number of iterations or calibration.  Of the seven parameter subsets, the optimal phytoplankton parameters had the largest reduction in \ac{rmse} values from the default output (starting \ac{rmse} $21.14$ mmol O$_2$ m$^{-3}$, reduction to $5.81$).  The identifiable parameters selected equally within categories (third heuristic) and independent of category (second heuristic) produced the second and third greatest reduction in \ac{rmse} ($19.09$, and $19.45$, respectively), although much less than that observed for phytoplankton parametrs. Overall, calibration to most of the parameter subsets had a small effect on \ac{rmse}.  However, the magnitude of parameter changes for calibration varied up to several orders of magnitude from the initial values (\cref{tab:calibprm}).  The percent change in parameters from the intial values varied from no change (Optics parameters \textit{astarOMZ} and ) to 900\% (\textit{KO2}).  Many calibrated parameter values included those at the lower or upper bounds used for the optimization routine (as for phytoplankton, \cref{fig:calex}).  There was no relationship between magnitude of parameter changes within a subset and overall reduction of \ac{rmse}.

The estimated areal extent of bottom-water hypoxia using default parameter values for the 3-dimensional model peaked on August, 28\textsuperscript{th} at $1.19e^{4}$ km$^2$ (\cref{fig:areachg}a, August mean area 8988 $\pm$ 551 km$^2$, \cref{fig:areachg}b). Increases in the top sensitive parameters for \ac{do} in each category (\cref{tab:dosens}) increased the estimated areal extent of hypoxia (Figs. \ref{fig:areachg}c, \ref{fig:areachg}e), with the exception of the parameter for optimum temperature of growth (\textit{Tref(nospA+nospZ)$_{p1}$}), which caused a decrease (mean difference from July default area -1070 $\pm$ 1161 km$^2$, \cref{fig:areachg}d).  The largest increase in area was caused by an increase in the maximum growth rate of phytoplankton (\textit{umax}), with the greatest increase occurring in July (monthly mean increase of 4109 $\pm$ 1085 km$^2$, \cref{fig:areachg}d).  Overall, changes in hypoxia extent were largest for parameters with large sensitivity values (i.e., L1 = $0.05$ for \textit{ZKa}, L1 = $0.05$ for \textit{umax}).

@@

Results from the identifiability analyses provided additional insight into the interactions of parameters in large biogeochemical models.  First, identifiability of parameter subsets was not related to the sensitivity of individual variables. As noted above, an identifiable parameter is one that has a unique effect on model predictions that cannot be compensated for or undone by changing other parameters. The magnitude of the effect of a parameter has no bearing on identifiability, which further complicates the selection of parameters for calibration.  Although identifiability is the primary limiting factor in choosing a set, the relative sensitivities are more important for the decision to include or exclude individual parameters.  For example, \cref{tab:calib,fig:areachg} demonstrated that parameter sets with sensitive parameters had improved model fit and that these parameters had affects that translated to the larger model. Our analysis addressed this challenge by presenting multiple selection criteria for identifiable parameter sets that prioritized the most sensitive parameters during the selection process.  Similarly, identifiability was not always related to the number of parameters in a set. Although the general trend was decreasing identifiability with more parameters, the unique effects of including an individual parameter with an existing set often reduced the $\gamma$ estimate. For example, \cref{fig:exclex} showed that including \textit{KGcdom}, \textit{KO2}, or \textit{nitmax} in parameter sets more often reduced $\gamma$ relative to sets that excluded the parameters.  The selection criteria proposed above can facilitate parameter selection and also provide diagnostic tools to identify parameters with disproportionate effects on $\gamma$.

@@

This study was funded, reviewed, and approved for publication by the US EPA, National Health and Environmental Effects Research Laboratory.  However, the views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA.

@@

\caption{Results of parameter changes from model calibration to observed \ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The parameters of each subset are from \cref{tab:heurist1,tab:heurist2,tab:heurist3}.  The L1 values are relative sensitivity of \ac{do} to a paramter change.  Starting and final parameter values after calibration are shown, including the percent change.  The calibration routing was bounded by minimum and maximum values appropriate for each parameter.\label{tab:calibprm}} 

@@

% model schematic
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{figs/fishmod.pdf}
\caption{Conceptual representation of biogeochemical components included in FishTank (complete equations in \citealt{Eldridge10}). The \acl{cgem} couples FishTank with a hydrodynamic model that includes advection, mixing, and dispersion.}
\label{fig:fishmod}
\end{figure}

@@

\caption{Example of calibration steps for the phytoplankton parameter subset.  The FishTank model was calibrated using data from an approximate 12-hour flask experiment of oxygen production.  Model calibration adjusted parameters (left plots) to minimize the \ac{rmse} between experimental and modelled dissolved oxygen (top left).  The phytoplankton parameters were those identifiable for dissolved oxygen (\cref{tab:heurist1}), arranged from most (top plot) to least sensitive (bottom plot) (\cref{tab:dosens}).  The relative parameter changes in the right plots show the initial parameter values (black) and the the selected parameter values (blue) after calibration.}\label{fig:calex}
\end{figure}

>>>>

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Thu Aug 17 18:08:14 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/7e3894b946e67efce7a075d46c2169be58729981}{7e3894b946e67efce7a075d46c2169be58729981}\newline Acronyms: \acfi{chla}; \acfi{cgem}; \acfi{do}; \acfi{dom}; \acfi{gom}; \acfi{lcs}; \acfi{marb}; \acfi{pom}; \acfi{rmse}; \acfi{zerod}}

@@

Estimates of oxygen production from closed bottle experiments were used to calibrate the \ac{zerod} model.  Surface water was collected from a fixed location in Pensacola Bay, Florida (30.35$^{\circ}$ N, 87.20$^{\circ}$ W) on September 25\textsuperscript{th}, 2013 (full details in \citetalias{MurrellIR}).  Water samples were placed into closed 300 mL glass bottles and transported back to the lab for treatments.  Concentrations of \ac{do} and nutrients (NH$_4^+$, NO$_2^-$, NO$_3^-$, PO$_4^{3-}$, and SiO$_3^{2-}$) were extracted from the samples within one hour of collection. Oxygen concentrations were established using Winkler reagents \citep{Parsons85} and titration using a MetrOhm Titrando with thiosulfate titrant calibrated against an iodate standard and electrochemical endpoint detection.  Ammonium was analyzed using a fluorometric method \citep{Holmes99}; other nutrients were analyzed using either a Thermo Fisher Aquakem 200 discrete analyzer or an Astoria-Pacific continuous flow analyzer using standard colorimetric methods \citepalias{APHA05}.

Oxygen production after an approximate 12 hour period (morning to evening) was estimated within the bottles using light and dark treaments.  All treatments were exposed to ambient outdoor conditions in a water bath at the laboratory.  The dark treatments used opaque black bottles to ensure no exposure of ambient light to the samples.  The exposure experiments began at 7am and were concluded at 7pm, after which the concentrations of \ac{do} in each bottle were measured.  Each treatment had four replicates and HOBO\textsuperscript{\textregistered} pendants were used to measure continuous temperature and light. Starting \ac{do} concentrations for the experiments were $200.78$ ($\pm$ $4.47$) mmol O$_2$ m$^{-3}$ and ending concentrations were $229.57$ ($\pm$ $2.66$) mmol O$_2$ m$^{-3}$ for the light treatments and $195.51$ ($\pm$ $1.9$) mmol O$_2$ m$^{-3}$ for the dark treatments.  Initial nutrient concentrations were $0.71$ mmol m$^{-3}$ for NH$_4^+$ (single sample), $0.13$ mmol m$^{-3}$ for PO$_4^{3-}$, and $38.4$ mmol m$^{-3}$ for SiO$_3^{2-}$. Concentrations for NO$_2^-$ and NO$_3^-$ were below detection limit.

Each parameter subset for O$_2$ from the selection heuristics was calibrated to the ending oxygen concentration for each of the four replicates of each treatment. The intial nutrient and oxygen concentrations were used as starting conditions for FishTank.  The corresponding continuous temperature and light measurements were also used as forcing conditions.  Parameter values were searched using the optim function in R that implemented the limited memory modification of the BFGS quasi-Newton method \citep{Byrd95,Nocedal06,RDCT17}.  This optimization scheme searches for values within the user-defined lower and upper bounds of each parameter.  We set the algorithm to search within a range of values for a subset that was chosen to best reflect expectations given professional judgment and reasonable bounds from the literature.  While we recognize that `actual' parameter values could possibly extend beyond these ranges, constraining the values was expected to produce output that was more interpretable within the context of the sensitivity functions defined above.  Moreover, precise limits on each parameter were less of a concern for the current analysis as our primary goal was to demonstrate differences in fit as a function of sensitivity and identifiability.  The optimal values for each parameter were identified based on convergence of the model results as a minimization of the \ac{rmse} between the observed and modelled data.

@@

An evaluation of the effects of individual parameters on $\gamma$ suggested that some parameters had disproportionate effects on identifiability.  For example, \cref{fig:identplo} suggests that most parameter sets for organic matter were identifiable ($\gamma <$ 15 for most), regardless of how many parameters were selected (i.e., two through eight).  However, some subsets were not identifiable such that identification of one or more redundant parameters that are inflating $\gamma$ values could provide useful information.  \Cref{fig:exclex} shows an alternative view of identifiability of \ac{do} with exclusion and inclusion of individual parameters in different sets for the organic matter category.  As before, collinearity increases with more parameters in a subset, although the increase varies depending on which parameter was included or excluded from the set.  For example, inclusion of \textit{KNO3} in a parameter set almost always inflated $\gamma$.  All parameter subsets that did not include \textit{KNO3} were well below $\gamma = 15$, suggesting that exclusion of this parameter improves identifiability.  Interestingly, the inclusion of some parameters caused a reduction in $\gamma$, which contradicts the general rule that more parameters caused reduced identifiablity.  For example, parameter sets that included \textit{KGcdom} generally had lower $\gamma$ values relative to those that excluded the parameter.

@@

Calibration of parameter values for \ac{do} for each of the seven selection heuristics (\cref{tab:heurist1,tab:heurist2,tab:heurist3}) produced model output that varied in goodness of fit and computational effort required to identify optimal values (\cref{tab:calib}).  As expected, the number of required iterations of the optimization algorithm to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only 12 iterations were required to optimize the single parameter selected for the temperature category, whereas 615 iterations were required for the seven parameters selected for phytoplankton (\cref{fig:calex}).  No association was observed between collinearity ($\gamma$) and number of iterations or calibration.  Of the seven parameter subsets, the optimal phytoplankton parameters had the largest reduction in \ac{rmse} values from the initial output (starting \ac{rmse} $21.14$ mmol O$_2$ m$^{-3}$, reduction to $5.81$).  The identifiable parameters selected equally within categories (third heuristic) and independent of category (second heuristic) produced the second and third greatest reduction in \ac{rmse} ($19.09$, and $19.45$, respectively), although much less than that observed for phytoplankton parameters. Overall, calibration to most of the parameter subsets had a small effect on \ac{rmse}.  However, the magnitude of parameter changes for calibration varied up to several orders from the initial values (\cref{tab:calibprm}).  The percent change in parameters from the intial values varied from no change (optics parameters \textit{astarOMZ} and \textit{astarOMA}) to 900\% (organic matter parameter \textit{KO2}).  Many calibrated parameter values included those at the lower or upper bounds used for the optimization routine (as for phytoplankton, \cref{fig:calex}).  There was no relationship between magnitude of parameter changes within a subset and overall reduction of \ac{rmse}.

The estimated areal extent of bottom-water hypoxia using initial parameter values for the 3-dimensional model peaked on August, 28\textsuperscript{th} at $1.19e^{4}$ km$^2$ (\cref{fig:areachg}a, August mean area 8988 $\pm$ 551 km$^2$, \cref{fig:areachg}b). Increases in the top sensitive parameters for \ac{do} in each category (\cref{tab:dosens}) increased the estimated areal extent of hypoxia (Figs. \ref{fig:areachg}c, \ref{fig:areachg}e), with the exception of the parameter for optimum temperature of growth (\textit{Tref(nospA+nospZ)$_{p1}$}), which caused a decrease (mean difference from July default area -1070 $\pm$ 1161 km$^2$, \cref{fig:areachg}d).  The largest increase in area was caused by an increase in the maximum growth rate of phytoplankton (\textit{umax}), with the greatest increase occurring in July (monthly mean increase of 4109 $\pm$ 1085 km$^2$, \cref{fig:areachg}d).  Overall, changes in hypoxia extent were largest for parameters with large sensitivity values (i.e., L1 = $0.05$ for \textit{ZKa}, L1 = $0.05$ for \textit{umax}).

@@

Results from the identifiability analyses provided additional insight into the interactions of parameters in large biogeochemical models.  First, identifiability of parameter subsets was not related to the sensitivity of individual variables. As noted above, an identifiable parameter is one that has a unique effect on model predictions that cannot be compensated for or undone by changing other parameters. The magnitude of the effect of a parameter has no bearing on identifiability, which further complicates the selection of parameters for calibration.  Although identifiability is the primary limiting factor in choosing a set, the relative sensitivities are more important for the decision to include or exclude individual parameters.  For example, \cref{tab:calib,fig:areachg} demonstrated that parameter sets with sensitive parameters had improved model fit and that these parameters had effects that translated to the larger model. Our analysis addressed this challenge by presenting multiple selection criteria for identifiable parameter sets that prioritized the most sensitive parameters during the selection process.  Similarly, identifiability was not always related to the number of parameters in a set. Although the general trend was decreasing identifiability with more parameters, the unique effects of including an individual parameter with an existing set often reduced the $\gamma$ estimate. For example, \cref{fig:exclex} showed that including \textit{KGcdom}, \textit{KO2}, or \textit{nitmax} in parameter sets more often reduced $\gamma$ relative to sets that excluded the parameters.  The selection criteria proposed above can facilitate parameter selection and also provide diagnostic tools to identify parameters with disproportionate effects on $\gamma$.

@@

This study was funded, reviewed, and approved for publication by the US EPA, National Health and Environmental Effects Research Laboratory.  However, the views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA. Mention of trade names or commercial products does not constitute endorsement by the authors or the US EPA.

@@

\caption{Results of parameter changes from model calibration to observed \ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The parameters of each subset are from \cref{tab:heurist1,tab:heurist2,tab:heurist3}.  The L1 values are relative sensitivity of \ac{do} to a parameter change.  Initial and final parameter values after calibration are shown, including the percent change.  The calibration routine was bounded by minimum and maximum values for each parameter.\label{tab:calibprm}} 

@@

% model schematic
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{figs/fishmod.pdf}
\caption{Conceptual representation of biogeochemical components included in FishTank (complete equations in \citealt{Eldridge10}, appendices A-F in \citealt{Lehrter17}). The \acl{cgem} couples FishTank with a hydrodynamic model that includes advection, mixing, and dispersion.}
\label{fig:fishmod}
\end{figure}

@@

\caption{Example of calibration steps for the phytoplankton parameter subset.  The FishTank model was calibrated using data from an approximate 12-hour flask experiment of oxygen production.  Model calibration adjusted parameters (left plots) to minimize the \ac{rmse} between experimental and modelled dissolved oxygen (top left).  The phytoplankton parameters were those identifiable for dissolved oxygen (\cref{tab:heurist1}), arranged from most (top plot) to least sensitive (bottom plot) (\cref{tab:dosens}).  The relative parameter changes in the right plots show the initial parameter values (black) and the selected parameter values (blue) after calibration.}\label{fig:calex}
\end{figure}
